Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance

### 2014-03-04 HRUIofMovingObjects/jrnl_article.tex
Reade for second submit.

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[draftcls]{IEEEtran}
%\documentclass[journal]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

@@

\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}. The method has been shown to improve lateral resolution and edge sharpness in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}. This is accomplished by adaptively adjusting the array weights based on element-to-element covariance in order to minimize interference. The method is, however, also highly sensitive to model errors \cite{Mestre2006, Widrow1982, Wax1996, Wax1996a}. This has led to the development of several robust versions of the Capon beamformer with e.g. spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets \cite{Lorenz2005, Rubsamen2013}. Spatial smoothing and diagonal loading, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}. The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution parameters \cite{Synnevag2007a}. Steering vector uncertainty sets, however, have not been applied for medical ultrasound imaging.

@@

In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots) \cite{Hergum2007}. A LSV-plot is constructed for a given imaging system by imaging a point scatterer moving laterally at constant speed. For each lateral position the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked in to an image. %For a given scan sequence there are twice as many point positions than there are receive lines. 
The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB. The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot. The LSV-plots based on simulations has been up-interpolated to 128 samples in each direction using cubic interpolation if the number of receive beams is less than 128.

@@

Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}). Dotted vertical lines indicate position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} using a 96 element phased array with a 22 mm azimuth aperture.  The probe was excited with a 2.5 MHz and 1.5 cycles long pulse. The transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point. The transmit focus is where the beam is at its narrowest, hence shift variance effects will appear at this location first. The beam density was calculated  based on (\ref{eq:resolution}) with the speed of sound equal to 1540. We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE). We therefore conclude that the system is locally laterally shift-invariant. %In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.

@@

\section{Local Lateral Shift-Invariance of the Capon Beamformer}\label{sec:capon_LLSI}
We will now investigate the local lateral shift-invariance of the Capon beamformer with high resolution parameters \cite{Synnevag2009, Asen}. In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 48$, $K=1$ and $d=1/100$. We see how the signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector increases as the point moves away from a given transmit-receive beam. As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig\,\ref{fig:das_capon_beams}. %The difference between the predicted and the simulated loss might be caused by several factors. The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS operator used in Fig.\,\ref{fig:capon1} average a broadband signal in range. For a slice through the center of the pulse the scalloping loss is below -30dB.

@@

From Fig.\,\ref{fig:capon_oversampling_b} we see that sixteen times oversampling reduces the scalloping loss significantly. Even though high frequency variations are reduced, the maximum is still larger than 1 dB. An even higher level of oversampling is therefore required in order to have a completely local lateral shift-invariant Capon beamformer when a moving point scatterer is simulated without noise and with the same parameters as in Fig\,\ref{fig:capon}. %$1/100$ in diagonal loading. 
Note that a diagonal loading factor of $1/100$ will introduce a white noise component in the sample covariance matrix $20$ dB lower than the signal present. Hence, the covariance has been estimated from data which appeared to have an SNR of 20 dB.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig4a.eps}\label{fig:capon_oversampling_a}
}
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig4b.eps}\label{fig:capon_oversampling_b}
}
}
\caption{Capon beamforming as described in Fig\,\ref{fig:capon} with oversampling on transmit. The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation. a) Four times oversampling ($q=4$). b) Sixteen times oversampling ($q=16$). The maximum scalloping loss is reduced, but it is still too high for the system to be locally laterally shift invariant.}
\label{fig:capon_oversampling}
\end{figure*}

@@

\subsection{Oversampling by Phase Rotation}
The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation. The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since ultrasound data have to be pre-delayed. However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors. In a narrow band setting the beam can make a sweep across the whole imaging sector through phase steering. This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius} and its references). The maximal steering angle is approximately given by
\begin{align}
\theta_{max} = \alpha q \delta\theta,
\end{align}
where $\alpha$ is the pulse length in $\lambda$'s, and $q$ and $\delta\theta$ is the oversampling factor and beam spacing respectively as defined in (\ref{eq:resolution}). For our simulations $\theta_{max} = 3\delta\theta$.
It is therefore possible to make a sweep based on phase rotation between a given beam and half-way to its two nearest neighbours without introducing large errors. Phase rotation is actually what the Capon beamformer uses internally to micro-steer away from bright point scatterers and to enhance edges.

For a uniform linear array located along the x-axis the steering vector $\vec{a}_\theta$ should be calculated as 
\begin{align}
\vec{a}_\theta = 
\begin{bmatrix}
e^{-j\frac{2\pi}{\lambda_c}x_0\sin(\theta)} \\
e^{-j\frac{2\pi}{\lambda_c}x_1\sin(\theta)} \\
\vdots \\ 
e^{-j\frac{2\pi}{\lambda_c}x_{L-1}\sin(\theta)}
\end{bmatrix}
\end{align}
where $x_i$ is the element position and $\theta$ is swept from $-\delta\theta/2$ to $\delta\theta/2$. Note that the steering vector must be calculated for an $L$ element symmetric subarray, even though the full array has $M$ elements. This subarray should have the same element pitch, $x_i - x_{i-1}$, as the full array. %The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \delta\theta/2$.

@@

Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig\,\ref{fig:capon} with four and sixteen times oversampling on receive respectively ($q$-factor). Note that the figure is zoomed compared with Fig\,\ref{fig:capon}. As with oversampling on transmit there is some variation when the point moves in and out of focus. The maximum scalloping loss is therefore large. The extreme local variations are however reduced significantly. Comparing Fig\,\ref{fig:capon_oversampling} with Fig\,\ref{fig:capon_oversampling_pr}, the errors introduced by the phase rotation method are best seen in sidelobes.

@@

\begin{figure*}[!t]
	\centerline{
		\subfloat[]{
			\includegraphics[width=0.5\linewidth]{\img fig5a.eps}
		}
		\subfloat[]{
			\includegraphics[width=0.5\linewidth]{\img fig5b.eps}
		}
	}
	\caption{Capon beamforming as described in Fig\,\ref{fig:capon} with oversampling based on phase rotation. The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation. a) Four times oversampling ($q=4$). b) Sixteen times oversampling ($q=16$).}
	\label{fig:capon_oversampling_pr}
\end{figure*}

\section{In-vitro phantom data}\label{sec:res}
In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation. The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S-D phased transducer operating in harmonic mode at 3.34 MHz. The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency, and 111 tx-beams where therefore distributed on a 65\degree sector. Transmit focus is located at 8 cm range. Note that in harmonic imaging the generated transmit beam at the harmonic frequency will be wider and have lower sidelobes than the beam resulting from transmitting at the harmonic frequency \cite{Fedewa2004}. The lateral sampling is therefore increased compared with our simulations. A tissue mimicking phantom containing wire targets and anechoic cysts was imaged, and the data was acquired while the probe was moved across the phantom manually. Fig.\,\ref{fig:image_das} shows the first frame out of 14 in the acquired loop processed with DAS beamforming. %Movies of the whole loop processed with DAS and Capon beamforming with and without ten times oversampling by phase rotation can be found in the supplementary material \multimedia{Videos}.

One point scatterer in Fig\,\ref{fig:image_das} has been selected [white box], and then used to generate LSV-plots. The point scatterer is located close to the transmit focus. Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer. The scalloping loss is less than 1 dB, and the system is by that locally laterally shift-invariant. As a consequence point scatterers have a consistent appearance when subjected to motion \multimedia{Media Movie 1}.

@@

Figure \ref{fig:phantom_capon_LSV_10PR} presents the result of applying sixteen times oversampling by phase rotation. We observe how the amplitude of the point scatterer is better preserved in all frames, and the scalloping loss is not far from our 1 dB threshold. Note that the oscillation along the diagonal does not indicate shift-variance. It is a result of lack of control when the probe was manually dragged across the phantom. In order to get a diagonal without peaks the probe has to be dragged across the phantom with a speed which matches the acquisition frame rate. In addition, due to the increase in lateral resolution, the Capon beamformer requires smaller frame-to-frame movements than DAS. Looking at the video of Capon beamforming with sixteen times oversampling by phase rotation \multimedia{Media Movie 3} the blinking is now highly reduced. %The system is however not fully shift invariant. To achieve this a higher level of oversampling, %We therefore conclude that the system is now locally laterally shift invariant.

@@

\begin{figure*}[!t]
\centerline{
\begin{tabular}{cc}
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6a.eps}\label{fig:image_das}}
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6b.eps}\label{fig:phantom_das}}\\
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6c.eps}\label{fig:phantom_capon_LSV}}
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6d.eps}\label{fig:phantom_capon_LSV_10PR}}
\end{tabular}
}
\caption{LSV-plots generated from \textit{in-vitro} phantom data where the probe has been manually dragged across the phantom. See the supplementary material for videos of the whole loop processed with the different beamformers \multimedia{Videos}. Capon beamforming has been performed with the same settings as in Fig.\,\ref{fig:capon1}. P-MAE is not shown due to a lack of probe movement control. a) First frame in the recorded loop processed with delay-and-sum beamforming. A white box indicates the point scatterer used to construct the LSV-plots.  b) Delay-and-sum beamforming. c) Capon beamforming without oversampling. d) Capon with sixteen times oversampling using phase rotation.}
\label{fig:LSV_invitro}
\end{figure*}

@@

Faced with an imaging scene containing bright point scatterers or sharp edges, one could wonder how large the oversampling factor needs to be in order to avoid visible scalloping losses. In his seminal paper, Cox \cite{Cox1973} derives the required beam density for a given output SNR. For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}) [with $qp=1$]. Since resolution for the Capon beamformer depends on SNR we would recommend to first use the dynamic range in the image as guidance. The amount of diagonal loading will also dictate the maximal achievable resolution since $10\log{d}$ can be interpreted as the maximal input SNR seen by the Capon beamformer. Fig\,\ref{fig:das_capon_beams} shows that 25 times oversampling is required for a $d$-value of $1/100$. When \textit{in-vitro} data contains additional noise this oversampling requirement is reduced.%Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves. For ultrasound data only a 50 \% increase in resolution was found in \cite[Fig.~7]{Asen} for a 30 dB output SNR when measured in a similar way as Cox [6 dB saddle point]. However, sixteen times oversampling was required to make the imaging system shift invariant in the same paper. In this paper more than ten times oversampling was found to be required for a 40 dB output SNR (after array gain) in simulations, however for the \textit{in-vitro} data ten times oversampling was enough for the system to be locally laterally shift-invariant. This might be a result of improved initial sampling (as discussed in Section \ref{sec:res}) and a higher initial noise level before diagonal loading was applied.

@@

Oversampling by phase rotation can therefore introduce similar block artifacts as parallel receive beamforming (PRB). To remove these artifacts we could distribute our steering angles in an overlapping scheme [like rx-beams are spread out for STB]. We then get a phase rotation version of STB where e.g. eight phase rotated beams are averaged into four. If the geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Media Movie 3} is removed by such technique is left for future work.

@@

%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
%(S12) was born in Porsgrunn, Norway in 1986. He received the B.Sc. and M.Sc. degree in computer science from the University of Oslo, Norway, in 2008 and 2010. He is currently pursuing his Ph.D. degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway.

%He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010. His research interests include image and signal processing, adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs). 
%\end{IEEEbiography}

%\vfill{}

@@

%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
%was born in Oslo, Norway, in 1970. He received the M.S. degree in physics in 1996 and the Ph.D. degree in computer science in 2001, both from the University of Oslo. Since 2001, he has been working at the Department of Informatics, University of Oslo, first as a postdoctoral research fellow and currently as an associate professor. His research interests include signal and array processing for acoustical imaging.
%\end{IEEEbiography}

>>>>

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
%\documentclass[draftcls]{IEEEtran}
\documentclass[journal]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

@@

\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer [also known as the minimum variance beamformer] for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}. The method has been shown to improve lateral resolution and edge sharpness in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}. This is accomplished by adaptively adjusting the array weights based on element-to-element covariance in order to minimize interference. The method is, however, also highly sensitive to model errors \cite{Mestre2006, Widrow1982, Wax1996, Wax1996a}. This has led to the development of several robust versions of the Capon beamformer with e.g. spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets \cite{Lorenz2005, Rubsamen2013}. Spatial smoothing and diagonal loading, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}. The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution parameters \cite{Synnevag2007a}. Steering vector uncertainty sets, however, have not been applied for medical ultrasound imaging.

@@

In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots) \cite{Hergum2007}. A LSV-plot is constructed for a given imaging system by imaging a point scatterer moving laterally at constant speed. For each lateral position the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked in to an image. %For a given scan sequence there are twice as many point positions than there are receive lines. 
The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB. The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot. The LSV-plots based on simulations have been up-interpolated to 128 samples in each direction using cubic interpolation if the number of receive beams is less than 128.

@@

Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}). Dotted vertical lines indicate position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} using a 96 element phased array with a 22 mm azimuth aperture, which was excited with a 2.5 MHz and 1.5 cycles long pulse. Transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point. The transmit focus is where the ultrasound beam is at its narrowest, hence shift variance effects will appear at this location first. The beam density was calculated  based on (\ref{eq:resolution}) with the speed of sound equal to 1540. We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE). We therefore conclude that the system is locally laterally shift-invariant. %In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.

@@

\section{Local Lateral Shift-Invariance of the Capon Beamformer}\label{sec:capon_LLSI}
We will now investigate the local lateral shift-invariance of the Capon beamformer with high resolution parameters \cite{Synnevag2009, Asen}. In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 48$, $K=1$ and $d=1/100$. We see how the signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector increases as the point moves away from a given transmit-receive beam. As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig.\,\ref{fig:das_capon_beams}. %The difference between the predicted and the simulated loss might be caused by several factors. The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS operator used in Fig.\,\ref{fig:capon1} average a broadband signal in range. For a slice through the center of the pulse the scalloping loss is below -30dB.

@@

From Fig.\,\ref{fig:capon_oversampling_b} we see that sixteen times oversampling reduces the scalloping loss significantly. Even though high frequency variations are reduced, the maximum is still larger than 1 dB. An even higher level of oversampling is therefore required in order to have a completely local lateral shift-invariant Capon beamformer when a moving point scatterer is simulated without noise and with the same parameters as in Fig.\,\ref{fig:capon}. %$1/100$ in diagonal loading. 
Note that a diagonal loading factor of $1/100$ will introduce a white noise component in the sample covariance matrix $20$ dB lower than the signal present. Hence, the covariance has been estimated from data which appeared to have an SNR of 20 dB.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig4a.eps}\label{fig:capon_oversampling_a}
}
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img fig4b.eps}\label{fig:capon_oversampling_b}
}
}
\caption{Capon beamforming as described in Fig.\,\ref{fig:capon} with oversampling on transmit. The figure is zoomed compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation. a) Four times oversampling ($q=4$). b) Sixteen times oversampling ($q=16$). The maximum scalloping loss is reduced, but it is still too high for the system to be locally laterally shift invariant.}
\label{fig:capon_oversampling}
\end{figure*}

@@

\subsection{Oversampling by Phase Rotation}
The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation. The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since ultrasound data have to be pre-delayed. However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors. In a narrow band setting the beam can make a sweep across the whole imaging sector through phase steering. This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius} and its references). The maximal steering angle is approximately given by
\begin{align}
\theta_{max} = 2\alpha \delta\theta,
\end{align}
where $\alpha$ is the pulse length in $\lambda$'s, and $2$ and $\delta\theta$ is the oversampling factor and beam spacing from (\ref{eq:resolution}). For our simulations $\theta_{max} = 3\delta\theta$.
It is therefore possible to make a sweep based on phase rotation between a given beam and half-way to its two nearest neighbours without introducing large errors. Phase rotation is actually what the Capon beamformer uses internally to micro-steer away from bright point scatterers and to enhance edges.

For a uniform linear array located along the x-axis the steering vector $\vec{a}_\theta$ should be calculated as 
\begin{align}
\vec{a}_\theta = 
\begin{bmatrix}
e^{-j\frac{2\pi}{\lambda_c}x_0\sin(\theta)} \\
e^{-j\frac{2\pi}{\lambda_c}x_1\sin(\theta)} \\
\vdots \\ 
e^{-j\frac{2\pi}{\lambda_c}x_{L-1}\sin(\theta)}
\end{bmatrix}
\end{align}
where $x_i$ is the element position and $\theta$ is swept from $-\delta\theta/2$ to $\delta\theta/2$. Note that the steering vector must be calculated for an $L$ element symmetric subarray, even though the full array has $M$ elements. This subarray should have the same element pitch, $|x_i - x_{i-1}|$, as the full array. %The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \delta\theta/2$.

@@

Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig.\,\ref{fig:capon} with four and sixteen times oversampling on receive respectively ($q$-factor). Note that the figure is zoomed compared with Fig.\,\ref{fig:capon}. As with oversampling on transmit there is some variation when the point moves in and out of focus. The maximum scalloping loss is therefore large. The extreme local variations are however reduced significantly. Comparing Fig.\,\ref{fig:capon_oversampling} with Fig.\,\ref{fig:capon_oversampling_pr}, the errors introduced by the phase rotation method are best seen in sidelobes.

@@

\begin{figure*}[!t]
	\centerline{
		\subfloat[]{
			\includegraphics[width=0.5\linewidth]{\img fig5a.eps}
		}
		\subfloat[]{
			\includegraphics[width=0.5\linewidth]{\img fig5b.eps}
		}
	}
	\caption{Capon beamforming as described in Fig.\,\ref{fig:capon} with oversampling based on phase rotation. The figure is zoomed compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation. a) Four times oversampling ($q=4$). b) Sixteen times oversampling ($q=16$).}
	\label{fig:capon_oversampling_pr}
\end{figure*}

\section{In-vitro phantom data}\label{sec:res}
In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation. The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S-D phased transducer operating in harmonic mode at 3.34 MHz. The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency, and 111 tx-beams where therefore distributed on a 65\degree sector. Transmit focus is located at 8 cm range. Note that in harmonic imaging the generated transmit beam at the harmonic frequency will be wider and have lower sidelobes than the beam resulting from transmitting at the harmonic frequency \cite{Fedewa2004}. The lateral sampling is therefore increased compared with our simulations. A tissue mimicking phantom containing wire targets and anechoic cysts was imaged, and the data was acquired while the probe was moved across the phantom manually. Fig.\,\ref{fig:phantom_das_image} shows the first frame out of 14 in the acquired loop processed with DAS beamforming. %Movies of the whole loop processed with DAS and Capon beamforming with and without ten times oversampling by phase rotation can be found in the supplementary material \multimedia{Videos}.

One point scatterer in Fig.\,\ref{fig:phantom_das_image} has been selected [white box], and then used to generate LSV-plots. The point scatterer is located close to the transmit focus. Figure \ref{fig:phantom_das_LSV} shows the resulting LSV-plot when using the DAS beamformer. The scalloping loss is less than 1 dB, and the system is by that locally laterally shift-invariant. As a consequence point scatterers have a consistent appearance when subjected to motion \multimedia{Media Movie 1}.

@@

Figure \ref{fig:phantom_capon_LSV_16PR} presents the result of applying sixteen times oversampling by phase rotation. We observe how the amplitude of the point scatterer is better preserved in all frames, and the scalloping loss is not far from our 1 dB threshold. Note that the oscillation along the diagonal does not indicate shift-variance. It is a result of lack of control when the probe was manually dragged across the phantom. In order to get a diagonal without peaks the probe has to be dragged across the phantom with a speed which matches the acquisition frame rate. In addition, due to the increase in lateral resolution, the Capon beamformer requires smaller frame-to-frame movements than DAS. Looking at the video of Capon beamforming with sixteen times oversampling by phase rotation \multimedia{Media Movie 3} the blinking is now highly reduced. The same can be seen from Fig.\,\ref{fig:img_invitro} where point scatterers appear brighter in the image with oversampling by phase rotation.%The system is however not fully shift invariant. To achieve this a higher level of oversampling, %We therefore conclude that the system is now locally laterally shift invariant.

@@

\begin{figure*}[!t]
\centerline{
\begin{tabular}{cc}
\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6a.png}\label{fig:phantom_das_image}}
\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6b.eps}\label{fig:phantom_das_LSV}}\\
\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6c.eps}\label{fig:phantom_capon_LSV}}
\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6d.eps}\label{fig:phantom_capon_LSV_16PR}}
\end{tabular}
}
\caption{LSV-plots generated from \textit{in-vitro} phantom data where the probe has been manually dragged across the phantom. See the supplementary material for videos of the whole loop processed with the different beamformers. Capon beamforming has been performed with the same settings as in Fig.\,\ref{fig:capon1}. P-MAE is not shown due to a lack of probe movement control. a) First frame in the recorded loop processed with delay-and-sum beamforming  \multimedia{Media Movie 1}. A white box indicates the point scatterer used to construct the LSV-plots.  b) LSV-plot for delay-and-sum beamforming. c) LSV-plot for Capon beamforming without oversampling. d) LSV-plot for Capon beamforming with sixteen times oversampling using phase rotation.}
\label{fig:LSV_invitro}
\end{figure*}

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=0.48\linewidth]{\img fig7a.png}\label{fig:phantom_capon_image}}
\subfloat[]{\includegraphics[width=0.48\linewidth]{\img fig7b.png}\label{fig:phantom_capon_16PR_image}}
}
\caption{First frame from Fig.\,\ref{fig:phantom_das_image} processed with Capon beamforming. a) Without oversampling \multimedia{Media Movie 2}. b) With sixteen times oversampling using phase rotation \multimedia{Media Movie 3}. Notice how the point scatterers appear brighter in the image with oversampling.}
\label{fig:img_invitro}
\end{figure*}

@@

Faced with an imaging scene containing bright point scatterers or sharp edges, one could wonder how large the oversampling factor needs to be in order to avoid visible scalloping losses. In his seminal paper, Cox \cite{Cox1973} derives the required beam density for a given output SNR. For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}) [with $q=1/2$]. Since resolution for the Capon beamformer depends on SNR we would recommend to first use the dynamic range in the image as guidance. The amount of diagonal loading will also dictate the maximal achievable resolution since $10\log{d}$ can be interpreted as the maximal input SNR seen by the Capon beamformer. Fig.\,\ref{fig:das_capon_beams} shows that 25 times oversampling is required for a $d$-value of $1/100$. When \textit{in-vitro} data contains additional noise this oversampling requirement is reduced.%Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves. For ultrasound data only a 50 \% increase in resolution was found in \cite[Fig.~7]{Asen} for a 30 dB output SNR when measured in a similar way as Cox [6 dB saddle point]. However, sixteen times oversampling was required to make the imaging system shift invariant in the same paper. In this paper more than ten times oversampling was found to be required for a 40 dB output SNR (after array gain) in simulations, however for the \textit{in-vitro} data ten times oversampling was enough for the system to be locally laterally shift-invariant. This might be a result of improved initial sampling (as discussed in Section \ref{sec:res}) and a higher initial noise level before diagonal loading was applied.

@@

Oversampling by phase rotation can therefore introduce similar block artifacts as parallel receive beamforming (PRB). To remove these artifacts we could distribute our steering vectors in an overlapping scheme [like rx-beams are spread out for STB]. We then get a phase rotation version of STB where e.g. eight phase rotated beams are averaged into four. If the geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Media Movie 3} is removed by such technique is left for future work.

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
(S12) was born in Porsgrunn, Norway in 1986. He received the B.Sc. and M.Sc. degree in computer science from the University of Oslo, Norway, in 2008 and 2010. He is currently pursuing his Ph.D. degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway.

He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010. His research interests include image and signal processing, adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs). 
\end{IEEEbiography}

\vfill{}

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
was born in Oslo, Norway, in 1970. He received the M.S. degree in physics in 1996 and the Ph.D. degree in computer science in 2001, both from the University of Oslo. Since 2001, he has been working at the Department of Informatics, University of Oslo, first as a postdoctoral research fellow and currently as an associate professor. His research interests include signal and array processing for acoustical imaging.
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
(M’82-SM’02) was born in Oslo, Norway, in 1954. He received the M.S. and Ph.D. degrees in electrical engineering from the Norwegian Institute of Technology, Trondheim, in 1978 and 1982, respectively. His academic experience includes Yarmouk University in Jordan (1984-1986) and the Norwegian Institute of Technology and later the University of Oslo (1989-1994) as an adjunct professor. Since 1995, he has been a professor of signal processing in the Department of Informatics at the University of Oslo. In 2002, he was elected a member of the Norwegian Academy of Technological Sciences, and in 2009, he became an adjunct professor in the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab, Trondheim. His industry experience includes GE Vingmed Ultrasound, Horten, Norway (1990-1994). He worked with Sonitor Technologies (2000-2005), where he developed ultrasonic technology that is the core of the indoor positioning system described in \textit{Scientific Americanin} 2008 as "a positioning system that goes where GPS can't." He has published about 130 publications internationally, plus about 10 patents. He has had sabbaticals at GE Global Research, NY (1998), and the Waves and Acoustics Laboratory of ESPCI, Paris (2008-2009). He was an associate editor of the \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control} from 1997 to 2002. His research interests are medical ultrasound imaging, modeling of tissue-ultrasound interaction, and ultrasound positioning.
\end{IEEEbiography}

\vfill
